Comparing methods for estimating parasite-induced host mortality from distributional data: implementation and limitations

### 2015-04-24 docs/parasite_mortatlity_manuscript.tex
Sent draft off to Sara

@@

Infectious agents [parasites?] can have major impacts on animal populations
through changing population dynamics and stability \citep{Dobson1992}, altering
predator prey interactions \citep{Joly2004}, and even causing species decline
and extinction \citep{DeCastro2005a,McCallum2012b}. Accurate estimates of
parasite-induced host mortality (PIHM) in wild animals are important for
understanding what regulates both host and parasite populations and to make
predictions about disease transmission in natural systems. Although a negative
impact on host fitness is a fundamental component of parasitism
\citep{Lafferty2002} [more], it is notoriously difficult to quantify PIHM in
wild animal populations \citep{Lester1984,McCallum2000a}.  While both macro and
microparasites have detrimental and possibly fatal effects on hosts, for the
remainder of this paper we limit our discussion to macroparasites that can be
discretely counted within a host.

[More discussion of PIHM before the nitty-gritty?]

To conclusively identify PIHM in wild animal populations, it is necessary to
experimentally infect and track host populations; a method that is rarely
possible in most wild animal systems \citep{McCallum2000a}.  Instead, parasitologists
are often only able to determine the parasite intensity on some number of sampled hosts.  This snapshot, distributional data is far from the
ideal type of data for addressing questions regarding PIHM, but this is the type of data on which most questions regarding PIHM are asked \citep[e.g.][]{Ferguson2011,Royce1990,Lanciani1989,Lester1984,Lester1977}.  In particular, parasitologists are often interested in asking two questions from this data: 1) Is PIHM occurring in this system? and if PIHM is occurring 2) Can we quantify the way in which parasite intensity is affecting host survival?

The first of these PIHM questions was addressed by \cite{Crofton1971a} who developed
a method to test for the presence of PIHM using the truncation of the negative binomial
distribution.  In short, the Crofton Method assumes that the distribution of
parasites across hosts before mortality occurs follows a negative binomial
distribution \citep{AndersonandMay1978,Shaw1998}.  As heavily infected hosts
begin to die, the negative binomial distribution is truncated and these hosts are no longer observed in a sample. In other words, the
observed host-parasite distribution and the pre-mortality host-parasite
distribution will predict substantially different numbers of of highly infected
hosts (because those have died due to infection), but similar number of hosts with low infection
loads (because those have survived).  \cite{Crofton1971a} noted that by
starting with all the observed data and iteratively fitting a negative binomial
distribution to hosts with lower and lower parasite loads, one could determine
whether or not PIHM was occurring in the system and estimate the parameters of the pre-mortality distribution.  This was done
graphically by determining whether parameters of the truncated negative binomial distributions showed a substantial change as hosts with smaller and smaller parasite intensities were fitted (Figure X).  The Crofton Method and this graphical technique are both still currently used \citep{Ferguson2011}. A more thorough description and implementation of the Crofton Method is given in Appendix X and we discuss the validity of its assumptions in the \emph{Discussion}.

The second question regarding PIHM attempts to quantify how parasite intensity affects host survival.  This question asks substantially more from the data and \cite{Adjei1986} proposed a method to attempt to answer it.  The Adjei Method proceeds
by first using the Crofton Method to estimate the pre-mortality parameters for
a host-parasite distribution and then, given these
parameters, estimates a host survival function that describes how the
probability of host-survival changes with increasing parasite load (see
\emph{Methods}).  From this host survival function, the
Adjei Method can estimate important host-parasite quantities such as the
parasite intensity at which 50\% of hosts succumb to PIHM ($LD_{50}$), as well
as the percent of hosts in a population succumbing to PIHM \citep{Adjei1986}. A more thorough description of the Adjei Method is given in Appendix X.

Despite these methods both being over three decades old, they are still the
primary means of answering questions about PIHM given distributional data
\citep[but see][for an alternative to the Crofton
Method]{Ferguson2011}. However, both methods have a few important limitations.
The Crofton Method, and more recent methods \citep{Ferguson2011},
rely on a visual test to determine whether or not PIHM is occurring in a
system.  With no clear decision rule, it can be difficult to determine the
significance of PIHM in a given host-parasite system and use consistent criteria for identifying PIHM across systems. In theory, the Adjei Method can ameliorate this problem. In addition to providing information on the host-survival function, this method can also be used to assess the significance of
PIHM in a system.
In practice, however, the Adjei Method has never been thoroughly tested and
relies on a number of questionable data manipulations.

In this study, we have
three primary goals. First we wish to test reliability of the Adjei Method for
answering the aforementioned questions regarding PIHM.  Second, we propose a
novel method for answering these questions and compares its efficacy against the Adjei Method.  Third, we
discuss the limitations of inferring PIHM from distributional data alone and
whether any method for inferring PIHM has a place in the future of
parasitology.

@@

The Adjei Method for estimating PIHM has two steps.  The first
step is to estimate the parameters of the pre-mortality host-parasite
distribution using the Crofton Method.  The three parameters estimated are the
total number of hosts before mortality $N_p$,  the mean number of parasites per
host before mortality $\mu_p$, and the aggregation of parasites before
mortality given by the parameter $k_p$ from a negative binomials distribution.
When $k_p$ is small, parasites are highly aggregated within hosts and when
$k_p$ is large parasites are more evenly spread among hosts \citep{Wilson2002}.  The implementation of the Crofton Method has been discussed at length elsewhere \citep[e.g.][]{Royce1990,Lester1984} and we provide a tested implementation of the method in Appendix X.

@@

\begin{equation}
    h(x | a, b) = h_x = \dfrac{e^{a - b \log(x)}}{1 + e^{a - b \log(x)}}
    \label{eq:logistic}
\end{equation}

where $x$ is the parasite intensity in a given host and $a$ and $b$ are the two
parameters of the function. Generally, a larger $a$ allows for hosts to
tolerate larger parasite loads before experiencing parasite-induced mortality
and a larger $b$ leads to a more rapid decline in the probability of host
survival as parasite intensity increases.  The value $\exp(a / b)$ is referred
to as the $LD_{50}$, which gives the parasite intensity at which 50\% of hosts
experience mortality.

To estimate this function the Adjei Method proceeds as follows.  First, it
calculates the expected number of hosts with a given parasite load $x$ by using
the equation $g(x ; \mu_p, k_p) * N_p$ where $g(x ; \cdot)$ is the negative binomial pre-
mortality distribution.  Second, the observed and predicted number of hosts
with $x$ parasites are paired as a single data point and the method then assumes that
this data point follows a binomial distribution with the total number of
``trials'' equal to the predicted number of hosts and the total number of
``successes'' equal to the observed number of hosts. In some cases, the
observed number of hosts is greater than the expected number of hosts and the
Adjei Method alters the data so that the observed is equal to the predicted
\citep{Adjei1986}.  After this questionable manipulation, the (observed, predicted) pairs are fit to a standard Generalized Linear Model \citep{McCullagh1989} with a binomial response variable and a logistic link function given by equation \ref{eq:logistic}.  This model provides estimates for parameters $a$, $b$ and $LD_{50}$.

@@

The Adjei Method's most glaring deficiency is the need to alter the observed
data in order to fit the model into the binomial GLM framework.  A second more
subtle problem with the Adjei Method is the potential need to bin data in order
to predict greater than one host in a given parasite intensity class.  For
example, if the total number of hosts pre-mortality was 50, the mean number of
parasites per host pre-mortality was 100 and the aggregation parameter was
$k_p$, applying the equation $g(x ; \mu_p=100, k_p=1) * 50$ would result in
less than 1 individual in all parasite intensities $x$. In other words, the
Adjei Method cannot be applied to samples with either very high mean parasite
loads, small sample sizes, or both without some sort of binning of the data.
While this is not a flaw ``per se'', it does add a certain level of
subjectivity (i.e. which bins should you use?) to a method that already has
serious potential issues.  In this analysis, we always assume the Adjei Method is not binning the data, though we provide code for applying the binning method in Appendix X.

@@

As with all previously proposed methods for estimating PIHM, the
Likelihood Method first assumes that the pre-mortality distribution follows a
negative binomial distribution $g(x; \mu_p, k_p)$.  The second
assumption that the likelihood method makes is that the host survival function
takes the form of a logistic curve given by equation \ref{eq:logistic}.  With these two explicit assumptions, the Likelihood Method estimates the 4 parameters $\mu_p$, $k_p$, $a$, and $b$.

@@

To estimate that significance of PIHM in a host-parasite system, can use a
likelihood ratio test in which the full model is given by equation
\ref{eq:dist} and the reduced model is given by a negative binomial
distribution.  If PIHM is significant in the system, the resulting likelihood
ratio statistic should be greater $\chi^2_{2,\text{ } \alpha=0.05}$.

\subsection*{Testing the ability of approaches to identify PIHM}

To test the ability of the Adjei Method and the Likelihood Method to identify whether or not PIHM was occurring in a system, we randomly generated data using the following procedure.  First, we drew $N_p$ randomly infected hosts from a
negative binomial distribution with parameters $\mu_p$ and $k_p$.  This represented the dataset observed before mortality. Second, we chose values of $a$ and $b$ and calculated the probability of survival
for all $N_p$ hosts using equation \ref{eq:logistic}.  Third, we drew $N_p$ random numbers from a uniform distribution
between 0 and 1 and if host survival probability was less than this random
number, the host experienced parasite-induced mortality.  The surviving
hosts comprised the dataset ($\mathbf{x}$) that would be obtained in the field, after PIHM.

Using both the pre-mortality and post-mortality simulated datasets,  we assumed
that the values of $N_p$, $\mu_p$, and $k_p$ were known and tested the ability of both methods to correctly determine whether or not PIHM was occurring.  While this scenario is unrealistic because the parameters $N_p$,
$\mu_p$, and $k_p$ are always unknown, we implemented this scenario as a baseline to
establish the efficacy of the methods independent of the estimates of $N_p$, $\mu_p$ and $k_p$.  For the Adjei Method, $N_p$, $\mu_p$, and $k_p$ are estimated using the Crofton Method, while $\mu_p$ and $k_p$ in the likelihood method can be estimated jointly with $a$ and $b$ or via the Crofton Method.   If a
method could not correctly predict whether or not PIHM was occurring under these idealistic conditions, we considered this strong evidence of the unreliability of this method.

We used three different values of $\mu_p$ (10, 50, 100) and for each $\mu_p$ we examined three different survival functions that had graduate, moderate, and sharp decreases in host survival with increasing parasite intensity.  For a given $\mu_p$, each survival function had the same $LD_{50}$, but different values of $a$ and $b$ (Table X).  We examined each $\mu_p$-survival function pair at  three levels of parasite
aggregation, $k_p = 0.1, 0.5, 1$ - realistic values of parasite aggregation in natural populations \citep{Shaw1998}.  For each of these parameter
combinations we simulated 150 datasets and tested the probability of each method incorrectly identifying PIHM in the pre-mortality dataset (type I error) and correctly identifying PIHM in the post-mortality dataset (power).  For each method, we used a likelihood ratio test to determine whether the full model with PIHM provided a significantly better fit than the reduced model without PIHM at significance level $\alpha = 0.05$.  We tested all each parameter combinations for pre-mortality population sizes of $N_p$ = [50, 100, 200, 300, 400, 500]. $N_p$ is not technically the sample size on which the methods are being
tested on the post-mortality data because PIHM reduces $N_p$ for each simulated
dataset.  We therefore used the average number of surviving hosts over all 150 simulations for a given parameter combination as our measure of sample size in the power simulations.

\subsection*{Testing ability of PIHM approaches to recover survival function}

To compare the ability of the Adjei Method and the Likelihood Method to recover $LD_{50}$ and the parameters $a$ and $b$ or the survival function, we used the same simulation procedure and parameter combinations described above.For each parameter
combination we simulated 150 datasets, estimated $a$, $b$, and $LD_{50}$ and calculated the standardized bias and
precision \citep{Walther2005} for these estimates over varying pre-mortality host population sizes  $N_p$ = [300, 500, 1000, 2000, 5000, 7500,
10000]. We used the average number of surviving hosts over all 150 simulations for a given parameter combination as our measure of sample size.

@@

We tested the ability of the Adjei Method and the Likelihood Method to identify
PIHM on 6 host-parasite datasets given in \cite{Crofton1971a} and 4 datasets
given in \cite{Adjei1986} (Table X). In the \cite{Crofton1971a} datasets, the host was
the snail \emph{Gammarus pulex} which acts as the intermediate host for the
acanthocephalan \emph{Polmorphus minutus}. In the \cite{Adjei1986} datasets,
the hosts were two species of lizard fish \emph{Saurida tumbil} and
\emph{Saurida undosquamis} that were infected by the cestode parasite
\emph{Callitetrarhynchus gracilis}.  Males and females of both fish species
were considered separately.

In both these studies, the authors reported PIHM and we test whether the Adjei
Method and/or the Likelihood Method also predicted PIHM. For the 6 datasets from
\cite{Crofton1971a}, we truncated the data at 4 parasites, applied the Crofton
Method to estimate the pre-mortality distribution, and then ran the Likelihood
Method and Adjei Method using these pre-mortality parameters.  For the
\cite{Adjei1986} datasets, we followed the same procedure as the authors and
first truncated the data at 2 parasites and then fit the Crofton Method for the
female fish of both species.  We then parameterized the male pre-mortality
distributions for each species with the results from the females.  Finally, we
applied the Adjei Method and the Likelihood Method to determine whether or not
PIHM was significant for these species and compared our results to those given by the authors.

@@

\subsection*{Detecting PIHM}

The Adjei Method showed highly inflated Type I error (i.e. falsely detected
PIHM where there was no PIHM) for all parameter combinations that we
considered.  This method also showed the unintuitive pattern of Type I error
rate decreasing as sample size decreased.  This pattern is due to the issue of
binning discussed in the \emph{Methods}. For small samples sizes, the
applicability of the Adjei Method is compromised without binning the observed
data in some way.  On the other hand, the Likelihood Method showed a type I
error rate at or near the pre-set level of 0.05 for all parameter combinations
and sample sizes considered.

The ability of the Adjei Method to detect PIHM given that it was occurring in a
system (i.e. the power) was close to 1 for larger sample sizes and tended to
decrease as sample size decreased.  The Likelihood Method was a power close to
one for all parameter combinations and sample sizes considered.  With gradual
survival functions (black lines; Figure X), the Likelihood Method showed a
decreased ability to detect PIHM for small sample sizes.

\subsection*{Recovery of the survival function}

The Likelihood Method gave asymptotically unbiased estimates of the $LD_{50}$
for all combinations of parameters examined in this study (Figure X).  Even
small sample sizes (< 500 hosts), the Likelihood Methods estimate of $LD_{50}$
was largely unbiased, with small biases occurring for host survival functions
that were gradual (i.e. $\mu_p = 10, a = 5, b = -2.5; \mu_p = 100, a = 5, b=
-1.5; \mu_p = 50, a = 5, b = -1.4$)   On the other hand, the Adjei Method
always produced more biased estimates of the $LD_{50}$ than the Likelihood
Method across all parameter combinations.  For $\mu_p = 10$, the $LD_{50}$
estimates from the Adjei Method were largely unbiased for large samples sizes,
but showed increasing bias as sample size decreased, particularly for steep
survival functions ($a = 20, b = -10$).  As $\mu_p$ increased, the Adjei method
produced biased estimates of $LD_{50}$ across all sample sizes, with bias
increasing as sample size decreased.

The precision of the $LD_{50}$ estimates for the Likelihood Method decreased
(increasing coefficient of variation) as sample size decreased for all
parameter combinations we examined.  The $LD_{50}$ estimates from the Adjei
Method showed a similar pattern, with large decreases in precision occurring
for the steepest survival function across all values of $\mu_p$ (Figure X).

In terms of the host survival function, the Likelihood Method gave
asymptotically unbiased estimates of $a$ and $b$ as sample size increased for
all parameter combinations considered.  However, as sample size decreased, the
Likelihood Method tended to produce severely biased estimates of $a$ and $b$.
This was generally more pronounced for steeper survival functions and more
aggregated pre-mortality distributions (Figure X).  The Adjei Method produced
biased estimates of $a$ and $b$ across all sample sizes, with the bias
consistently being larger when the survival function was steeper. The bias of
the Adjei Method also increased as $\mu_p$ increased.

@@

Of the 10 datasets we considered, the previous authors visually detect PIHM
in 7 of them (Table \ref{table:pihm}).  The Likelihood Method parameterized
from the pre-mortality parameters of the Crofton Method detected significant
PIHM in 6 of these 7 datasets at a significance level of 0.05.  The only
dataset in which the the Likelihood Method parameterized from the Crofton
Method did not detect a significant effect of PIHM was the the Adjei dataset
for \emph{S. tumbil}.  For this dataset there was a marginally significant effect
of PIHM ($\chi^2_{df=2} = 5.34; p = 0.069$).

The Adjei Method detected significant PIHM in 9 of the 10 datasets given (Table \ref{table:pihm}).  This is consistent with the the previous results which show that the Adjei Method has a very high Type I error rate.  The Likelihood Method in which all parameters were jointly estimated predicted significant PIHM in 4 of the 10 datasets.

@@

Determining whether PIHM is a significant factor for a host populations is critically important in many systems, but detecting and describing PIHM given only observational data is
notoriously difficult.  We show that the Adjei Method, the only currently
proposed method to estimate the host survival function and the $LD_{50}$ from
observational PIHM, has some serious methodological problems that result in
biased estimates of the host survival function and the $LD_{50}$ even under the most idealistic conditions.  Moreover, we show that the Adjei Method has a seriously inflated Type I error rate, meaning it will often detect PIHM even when it is not present. Moreover, for small, realistic sample sizes the Adjei Method behaves erratically; a consequence of the need to subjectively bin the data in order to predict parasite intensity classes with at least one host.

>>>>

\section*{Abstract}

TODO

@@

Infectious agents can have major impacts on animal populations
through changing population dynamics and stability \citep{Dobson1992}, altering
predator-prey interactions \citep{Joly2004}, and even causing species decline
and extinction \citep{DeCastro2005a,McCallum2012b}. Accurate estimates of
parasite-induced host mortality (PIHM) in wild animals are important for
understanding what regulates both host and parasite populations and to make
predictions about disease transmission in natural systems. Although a negative
impact on host fitness is a fundamental component of parasitism
\citep{Lafferty2002} [more], it is notoriously difficult to quantify PIHM in
wild animal populations \citep{Lester1984,McCallum2000a}.

To conclusively identify PIHM in wild animal populations, it is necessary to
experimentally infect and track host populations; a method that is rarely
possible in most wild animal systems \citep{McCallum2000a}.  Instead, parasitologists
are often only able to determine the parasite intensity on some number of sampled hosts.  This snapshot, distributional data is far from the
ideal type of data for addressing questions regarding PIHM, but this is the type of data on which most questions regarding PIHM are asked \citep{Ferguson2011,Royce1990,Lanciani1989,Lester1984,Lester1977}.  In particular, parasitologists are often interested in asking two questions from this data: 1) Is PIHM occurring in a host-parasite system? and if PIHM is occurring 2) Can the effect of parasite intensity on host survival be quantified? While both macro and
microparasites have detrimental and possibly fatal effects on hosts, for the
remainder of this paper we limit our discussion to macroparasites that can be
discretely counted within a host.

The first of these PIHM questions was addressed by \cite{Crofton1971a} who developed
a method to test for the presence of PIHM using the truncation of the negative binomial
distribution.  In short, the Crofton Method assumes that the distribution of
parasites across hosts before mortality occurs follows a negative binomial
distribution \citep{AndersonandMay1978,Shaw1998}.  As heavily infected hosts
begin to die, the negative binomial distribution is truncated and these hosts are no longer observed in a sample. In other words, the
observed host-parasite distribution and the pre-mortality host-parasite
distribution will predict substantially different numbers of hosts with high infect intensity (because those have died due to infection), but a similar number of hosts with low infection
intensity (because those have survived).  \cite{Crofton1971a} noted that by
starting with all observed hosts and iteratively fitting a negative binomial
distribution to hosts with lower and lower parasite loads, one could determine
whether or not PIHM was occurring and estimate the parameters of the host distribution before parasite-induced mortality.  This was done
graphically by determining whether the parameters of the negative binomial distributions fit to different truncations of the data showed a substantial change as the truncation point moved from heavily infected hosts to lightly infected hosts.  The Crofton Method and this graphical technique are both still currently used to assess whether PIHM is a occurring in a system \citep{Ferguson2011}. We give a thorough description and implementation of the Crofton Method in \emph{Supplementary Information} (SI) 1 and discuss the validity of its assumptions in the \emph{Discussion}.

The second question regarding PIHM moves beyond a simple yes or no answer and attempts to quantify how parasite intensity affects host survival. \cite{Adjei1986} proposed a method to answer this question
by first using the Crofton Method to estimate the pre-mortality parameters for
a host-parasite distribution and then, given these
parameters, estimating a host survival function that described how the
probability of host-survival changed with increasing parasite load (see
\emph{Methods}).  With this host survival function, one can estimate important host-parasite quantities such as the
parasite intensity at which 50\% of hosts succumb to PIHM ($LD_{50}$), as well
as the percent of hosts in a population succumbing to PIHM \citep{Adjei1986}.

Despite these methods both being over three decades old, they are still the
primary means of answering questions about PIHM given distributional data
\citep[but see][for an alternative to the Crofton
Method]{Ferguson2011}. However, both methods have a few important limitations.
The Crofton Method, and more recent methods \citep{Ferguson2011},
rely on a visual test to determine whether or not PIHM is occurring in a
system.  With no clear decision rule, it can be difficult to consistently determine the
significance of PIHM across different host-parasite systems. In theory, the Adjei Method can ameliorate this problem. In addition to providing information on the host-survival function, this method can also be used to assess the significance of
PIHM in a system.
In practice, however, the Adjei Method has never been thoroughly tested and
relies on a number of questionable data manipulations.

In this study, we have
three primary goals. First we wish to test reliability of the Adjei Method for
answering the aforementioned questions regarding PIHM.  Second, we propose a
novel method for answering these questions, compare its efficacy against the Adjei Method, and test both methods ability to detect PIHM on empirical data.  Third, we
discuss the limitations of inferring PIHM from distributional data alone and
whether any method for inferring PIHM has a place in the future of
parasitology.

@@

The Adjei Method for estimating PIHM has two steps.  The first
step is to estimate the parameters of the pre-mortality host-parasite
distribution using the Crofton Method.  The three parameters estimated are the
total number of hosts before mortality $N_p$,  the mean number of parasites per
host before mortality $\mu_p$, and the aggregation of parasites before
mortality given by the parameter $k_p$ from a negative binomial distribution.
When $k_p$ is small, parasites are highly aggregated among hosts and when
$k_p$ is large parasites are more evenly spread out \citep{Wilson2002}.  The implementation of the Crofton Method has been discussed at length elsewhere \citep[e.g.][]{Royce1990,Lester1984} and we provide a tested implementation of the method in SI 3.

@@

\begin{equation}
    h(x | a, b) = h_x = \dfrac{e^{a - b \log(x)}}{1 + e^{a - b \log(x)}}
    \label{eq:logistic}
\end{equation}
where $x$ is the parasite intensity in a given host and $a$ and $b$ are the two
parameters of the logistic function. Generally, a larger $a$ allows for hosts to
tolerate larger parasite intensities before experiencing parasite-induced mortality
and a larger $b$ leads to a more rapid decline in the probability of host
survival as parasite intensity increases.  The value $\exp(a / b)$ is referred
to as the $LD_{50}$, which gives the parasite intensity at which 50\% of hosts
experience mortality.

To estimate this function the Adjei Method proceeds as follows.  First, it
calculates the expected number of hosts with a given parasite load $x$ by using
the equation $g(x ; \mu_p, k_p) * N_p$ where $g(x ; \cdot)$ is the negative binomial pre-mortality distribution.  Second, the observed and predicted number of hosts
with $x$ parasites are paired as a single data point and the method then assumes that
this data point follows a binomial distribution with the total number of
``trials'' equal to the predicted number of hosts and the total number of
``successes'' equal to the observed number of hosts. In some cases, the
observed number of hosts is greater than the expected number of hosts and the
Adjei Method alters the data so that the observed is equal to the predicted
\citep{Adjei1986}.  After this questionable manipulation, the (observed, predicted) pairs are fit to a standard Generalized Linear Model \citep{McCullagh1989} with a binomial response variable and a logistic link function given by equation \ref{eq:logistic}.  This model provides estimates for parameters $a$, $b$ and $LD_{50}$.

@@

The Adjei Method's most glaring deficiency is the need to alter the observed
data in order to fit the model into the binomial GLM framework.  A second more
subtle problem with the Adjei Method is the potential need to bin data in order
to predict greater than one host in a given parasite intensity class.  For
example, if the total number of hosts pre-mortality was 50, the mean number of
parasites per host pre-mortality was 100 and the aggregation parameter was 1, applying the equation $g(x ; \mu_p=100, k_p=1) * 50$ would result in
less than 1 individual in all parasite intensities $x$. In other words, the
Adjei Method cannot be applied to samples with either very high mean parasite
loads, small sample sizes, or both without some sort of binning of the data.
While this is not a flaw \emph{per se}, it does add a certain level of
subjectivity (i.e. which bins should you use?) to a method that already has
serious potential issues.  In this analysis, we always assume the Adjei Method is not binning the data, though we provide code for applying the binning method in SI 3.

@@

As with all previously proposed methods for estimating PIHM, the
Likelihood Method first assumes that the pre-mortality distribution follows a
negative binomial distribution $g(x; \mu_p, k_p)$.  The second
assumption is that the host survival function
takes the form of a logistic curve given by equation \ref{eq:logistic}.  With these two explicit assumptions, the Likelihood Method estimates the 4 parameters $\mu_p$, $k_p$, $a$, and $b$.

@@

To estimate that significance of PIHM in a host-parasite system, a
likelihood ratio test can be used in which the full model is given by equation
\ref{eq:dist} and the reduced model is given by a negative binomial
distribution.  If PIHM is not significant in the system, the resulting likelihood
ratio statistic should approximately follow a $\chi^2$ distribution with degrees of freedom equal to 2.  We provide the code for implementing this Likelihood Method in SI 3.

\subsection*{Question 1: Is PIHM occurring?}

To test the ability of the Adjei Method and the Likelihood Method to identify whether or not PIHM was occurring in a system, we randomly generated data using the following procedure.  First, we drew $N_p$ randomly infected hosts from a
negative binomial distribution with parameters $\mu_p$ and $k_p$.  This represented the dataset observed before mortality. Second, we chose values of $a$ and $b$ and calculated the probability of survival
for all $N_p$ hosts using equation \ref{eq:logistic}.  Third, we drew $N_p$ random numbers from a uniform distribution
between 0 and 1 and if host survival probability was less than this random
number, the host experienced parasite-induced mortality.  The surviving
hosts comprised the dataset that would be obtained in the field, after PIHM.

Using both the pre-mortality and post-mortality simulated datasets,  we assumed
that the values of $N_p$, $\mu_p$, and $k_p$ were known and tested the ability of both methods to correctly determine whether or not PIHM was occurring.  While this scenario is unrealistic because the parameters $N_p$,
$\mu_p$, and $k_p$ are always unknown, we implemented this scenario as a baseline to
establish the efficacy of the methods independent of the estimates of $N_p$, $\mu_p$ and $k_p$.  For the Adjei Method, $N_p$, $\mu_p$, and $k_p$ are estimated using the Crofton Method, while $\mu_p$ and $k_p$ in the Likelihood Method can be estimated jointly with $a$ and $b$ or via the Crofton Method.   If a
method could not correctly predict whether or not PIHM was occurring under these idealistic conditions, we considered this strong evidence of the unreliability of this method.

We used three different values of $\mu_p$ (10, 50, 100) and for each $\mu_p$ we examined three different survival functions that had graduate, moderate, and steep decreases in host survival with increasing parasite intensity (Figure \ref{fig:surv}).  For a given $\mu_p$, each survival function had the same $LD_{50}$ ([$\mu_p = 10$, $LD_{50} = 7.39$], [$\mu_p = 50$, $LD_{50} = 35.57$], [$\mu_p = 100$, $LD_{50}= 77.3$]),  but different values of $a$ and $b$.  We examined each $\mu_p$-survival function pair at  three levels of parasite
aggregation, $k_p = 0.1$, 0.5, and 1 --- realistic values of parasite aggregation in natural populations \citep{Shaw1998}.  For each of these parameter
combinations we simulated 150 datasets and tested the probability of each method incorrectly identifying PIHM in the pre-mortality dataset (Type I error) and correctly identifying PIHM in the post-mortality dataset (power).  For each method, we used a likelihood ratio test to determine whether the full model with PIHM provided a significantly better fit than the reduced model without PIHM at significance level $\alpha = 0.05$.  We tested each parameter combinations for pre-mortality population sizes of $N_p$ = [50, 100, 200, 300, 400, 500]. $N_p$ is not technically the sample size on which the methods are being
tested on the post-mortality data because PIHM reduces $N_p$ for each simulated
dataset.  We therefore used the average number of surviving hosts over all 150 simulations for a given parameter combination as our measure of sample size in the power simulations.

\subsection*{Question 2: Can the effect of parasite intensity on host survival be quantified?}

To compare the ability of the Adjei Method and the Likelihood Method to recover $LD_{50}$ and the parameters $a$ and $b$ of the survival function, we used the same simulation procedure and parameter combinations described above.For each parameter
combination we simulated 150 datasets, estimated $a$, $b$, and $LD_{50}$ and calculated the standardized bias and
precision \citep{Walther2005} for these estimates over varying pre-mortality host population sizes  $N_p$ = [300, 500, 1000, 2000, 5000, 7500,
10000]. We used the average number of surviving hosts over all 150 simulations for a given parameter combination as our measure of sample size.

@@

We tested the ability of the Adjei Method and the Likelihood Method to identify
PIHM in 6 host-parasite datasets given in \cite{Crofton1971a} and 4 datasets
given in \cite{Adjei1986} (Table \ref{table:pihm}). In the \cite{Crofton1971a} datasets, the host was
the snail \emph{Gammarus pulex} which acts as the intermediate host for the
acanthocephalan \emph{Polmorphus minutus}. In the \cite{Adjei1986} datasets,
the hosts were two species of lizard fish \emph{Saurida tumbil} and
\emph{Saurida undosquamis} that were infected by the cestode parasite
\emph{Callitetrarhynchus gracilis}.  Males and females of both fish species
were considered separately.

In both these studies, the authors reported PIHM in some of the datasets and we test whether the Adjei
Method and/or the Likelihood Method also predicted PIHM. For the 6 datasets from
\cite{Crofton1971a}, we truncated the data at 4 parasites, applied the Crofton
Method to estimate the pre-mortality distribution, and then ran the Likelihood
Method and Adjei Method using these pre-mortality parameters.  For the
\cite{Adjei1986} datasets, we followed the same procedure as the authors and
first truncated the data at 2 parasites and then fit the Crofton Method for the
female fish of both species.  We then parameterized the male pre-mortality
distributions for each species with the results from the females.  Finally, we
applied the Adjei Method and the Likelihood Method to determine whether or not
PIHM was significant for these species and compared our results to those given by the authors.  All fitting to data was done with the Python code provided in SI 3.

@@

\subsection*{Question 1: Is PIHM occurring?}

The Adjei Method showed highly inflated Type I error rates (i.e. falsely detected
PIHM) for all parameter combinations that we
considered (Figure \ref{fig:typeI10}; SI2 Figs 1, 2).  This method also showed the unintuitive pattern of Type I error
rate decreasing as sample size decreased.  This pattern is due to the issue of
binning discussed in the \emph{Methods}. For small samples sizes, the
applicability of the Adjei Method is compromised without binning the observed
data in some way.  In contrast, the Likelihood Method showed a Type I
error rate at or near the pre-set level of 0.05 for all parameter combinations
and sample sizes considered.

The ability of the Adjei Method to detect PIHM given that it was occurring in a
system (i.e. the power) was close to 1 for larger sample sizes and tended to
decrease as sample size decreased (Figure \ref{fig:typeI10}; SI2 Figs 1, 2).  The Likelihood Method had a power close to
one for all parameter combinations and sample sizes considered.  With gradual
survival functions (solid black lines; Figure \ref{fig:typeI10}, SI2 Figs 1, 2), the Likelihood Method showed a
decreased power to detect PIHM for small sample sizes.

% TODO: Need to figure out what to do with the mu = 100 that are being dropped
% because the sample size is too small!

\subsection*{Question 2: Can the effect of parasite intensity on host survival be quantified?}

The Likelihood Method gave asymptotically unbiased estimates of the $LD_{50}$
for all combinations of parameters examined in this study (Figure \ref{fig:biasld50}, SI2 Fig 3, 4).  Even for
small sample sizes (< 500 hosts), the Likelihood Method's estimate of $LD_{50}$
was largely unbiased, with small biases occurring for host survival functions
that were gradual.  On the other hand, the Adjei Method
always produced more biased estimates of the $LD_{50}$ than the Likelihood
Method across all parameter combinations (Figure \ref{fig:biasld50}, SI2 Fig 3, 4).  For $\mu_p = 10$, the $LD_{50}$
estimates from the Adjei Method were largely unbiased for large samples sizes,
but showed increasing bias as sample size decreased, particularly for steep
survival functions.  As $\mu_p$ increased, the Adjei method
produced biased estimates of $LD_{50}$ across all sample sizes, with bias
increasing as sample size decreased (SI2 Fig 3, 4).

The precision of the $LD_{50}$ estimates for the Likelihood Method decreased
(increasing coefficient of variation) as sample size decreased for all
parameter combinations we examined (Figure \ref{fig:biasld50}, SI2 Fig 3, 4).  The $LD_{50}$ estimates from the Adjei
Method showed a similar pattern, with large decreases in precision occurring
for the steepest survival function across all values of $\mu_p$ (Figure \ref{fig:biasld50}, SI2 Fig 3, 4).

In terms of the host survival function, the Likelihood Method gave
asymptotically unbiased estimates of $a$ and $b$ as sample size increased for
all parameter combinations considered (Figure \ref{fig:biasa}, SI2 Fig 5, 6).  However, as sample size decreased, the
Likelihood Method tended to produce severely biased estimates of $a$ and $b$.
This was generally more pronounced for steeper survival functions and more
aggregated pre-mortality distributions (e.g. Figure \ref{fig:biasa}).  The Adjei Method produced
biased estimates of $a$ and $b$ across all sample sizes, with the bias
consistently being larger when the survival function was steeper. The bias of
the Adjei Method's estimate of $a$ also increased as $\mu_p$ increased (Figure \ref{fig:biasa}, SI2 Fig 5, 6).

@@

Of the 10 datasets we considered, the previous authors visually detected PIHM
in 7 of them (Table \ref{table:pihm}).  The Likelihood Method parameterized
from the pre-mortality parameters of the Crofton Method detected significant
PIHM in 6 of these 7 datasets at a significance level of 0.05.  The only
dataset in which the the Likelihood Method did not detect a significant effect of PIHM was the Adjei dataset
for female \emph{S. tumbil}.  For this dataset there was a marginally significant effect
of PIHM ($\chi^2_{df=2} = 5.34; p = 0.069$).

The Adjei Method detected significant PIHM in 9 of the 10 datasets given (Table \ref{table:pihm}).  This is consistent with the previous results which show that the Adjei Method has a very high Type I error rate.

@@

Determining whether PIHM is a significant factor for a host populations is
critically important in many systems, but detecting and describing PIHM given
only observational data is notoriously difficult.  We show that the Adjei
Method, the only currently proposed method to estimate the host survival
function and the $LD_{50}$ from observational data, has some serious
methodological problems that result in biased estimates even under the most idealistic conditions.
Moreover, we show that the Adjei Method has a seriously inflated Type I error
rate, meaning it will often detect PIHM even when it is not present. Moreover,
for small, realistic sample sizes the Adjei Method behaves erratically; a
consequence of the need to subjectively bin the data in order to predict
parasite intensity classes with at least one host.
